---
title: "WES: BC8 crude counts - all"
author: "Alexey Larionov"
date: "01 Feb 2021"
output:
  html_document:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Summary

- Select pthogenic variants in BC8, including consensus high-grade missenses
- Calculate crude counts: totals, and separately for cases and controls  
- Fisher test for the crude counts  

Check call rates (per variant and per sample) before and after filtering.  

<style>
pre{
  overflow-x: auto;
}
pre code{
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width=999)
```

# Start section

```{r}

# Time
Sys.time()

# Memory
gc()

# Clean up
rm(list=ls())
graphics.off()

# Options
options(stringsAsFactors = F)

# Working folders
base_folder <- "/Users/alexey/Documents"
project_folder <- file.path(base_folder,"wecare","final_analysis_2021","reanalysis_wo_danish_2021","s01_wes_wecare_only")
scripts_folder <- file.path(project_folder,"scripts","s08_crude_counts")
setwd(scripts_folder)
data_folder <- file.path(project_folder,"data","s08_crude_counts")

# Libraries
library(dplyr)
library(ggVennDiagram)

```

# Load data

```{r}

load(file.path(data_folder,"s01_crude_counts_per_variant.RData"))

# Update working folders
base_folder <- "/Users/alexey/Documents"
project_folder <- file.path(base_folder,"wecare","final_analysis_2021","reanalysis_wo_danish_2021","s01_wes_wecare_only")
scripts_folder <- file.path(project_folder,"scripts","s08_crude_counts")
data_folder <- file.path(project_folder,"data","s08_crude_counts")

bc8_genes <- scan(file.path(data_folder,"bc8_genes.txt"),what="character")

```

# Check data

```{r}

ls()

# Check objects dimentions
dim(genotypes.mx)
dim(phenotypes.df)
dim(variants.df)

# Check that the tables and matrices are in sync
sum(phenotypes.df$wes_id != colnames(genotypes.mx))
sum(variants.df$ID != rownames(genotypes.mx))

# Check list of genes
bc8_genes

```

# Select variants 

## Keep only variants in the selected genes

```{r}

bc8_variants.df <- variants.df[variants.df$vep_SYMBOL %in% bc8_genes,]
dim(bc8_variants.df)

```

## predicted_lof-s as defined by VEP IMPACT

```{r}

predicted_lof <- bc8_variants.df$vep_IMPACT == "HIGH"
sum(predicted_lof)

```

## Select known (likely) known_pathogenic by ClinVar

```{r}

# Explore ClinVar annotations
bc8_variants.df %>% 
  group_by(CLNSIG) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

# Select known_pathogenic variants
known_pathogenic <- 
  grepl("Pathogenic",bc8_variants.df$CLNSIG) | 
  grepl("Likely_pathogenic",bc8_variants.df$CLNSIG)
sum(known_pathogenic)

#grepl("risk_factor",bc8_variants.df$CLNSIG)

#bc8_variants.df %>% 
#  group_by(CLNREVSTAT) %>% 
#  summarise(counts=n()) %>% 
#  arrange(desc(counts))

#with_good_support <-
#  variants.df$CLNREVSTAT == "reviewed_by_expert_panel" |
#  variants.df$CLNREVSTAT == "criteria_provided,_multiple_submitters,_no_conflicts" |
#  variants.df$CLNREVSTAT == "criteria_provided,_single_submitter""
#etc

```

## Consensus high grade missenses  

```{r}

# Split SIFT
bc8_variants.df <- 
  bc8_variants.df %>% 
  mutate(SIFT_call=sub("\\(.*\\)","",vep_SIFT)) %>% 
  mutate(SIFT_score=as.numeric(
    sub(".*\\(","", sub("\\)","",vep_SIFT)))) %>% 
  select(-vep_SIFT)

bc8_variants.df %>% 
  group_by(SIFT_call) %>% 
  summarise(count=n()) %>% 
  arrange(desc(count))

# Split PolyPhen
bc8_variants.df <- 
  bc8_variants.df %>% 
  mutate(PolyPhen_call=sub("\\(.*\\)","",vep_PolyPhen)) %>% 
  mutate(PolyPhen_score=as.numeric(
    sub(".*\\(","", sub("\\)","",vep_PolyPhen)))) %>% 
  select(-vep_PolyPhen)

bc8_variants.df %>% 
  group_by(PolyPhen_call) %>% 
  summarise(count=n()) %>% 
  arrange(desc(count))

class(bc8_variants.df$vep_CADD_PHRED)
head(bc8_variants.df$vep_CADD_PHRED)
sum(is.na(bc8_variants.df$vep_CADD_PHRED))

# Consensus high grade missenses
consensus_high_grade_missenses <- 
  bc8_variants.df$SIFT_call == "deleterious" & !is.na(bc8_variants.df$SIFT_call) & 
  bc8_variants.df$PolyPhen_call == "probably_damaging" & !is.na(bc8_variants.df$PolyPhen_call) & 
  bc8_variants.df$vep_CADD_PHRED >=20 & !is.na(bc8_variants.df$vep_CADD_PHRED)

sum(consensus_high_grade_missenses)

```

## Rare variants

```{r}

# Rare in gnomeD NFE
sum(bc8_variants.df$vep_gnomAD_NFE_AF <= 0.01, na.rm=T)

min(bc8_variants.df$vep_gnomAD_NFE_AF, na.rm=T)
max(bc8_variants.df$vep_gnomAD_NFE_AF, na.rm=T) # no rare variants with MAF<0.01 and AF>0.99

sum(is.na(bc8_variants.df$vep_gnomAD_NFE_AF))

rare_in_gnomAD_NFE <- bc8_variants.df$vep_gnomAD_NFE_AF <= 0.01 | 
  is.na(bc8_variants.df$vep_gnomAD_NFE_AF)
sum(rare_in_gnomAD_NFE)

# Rare in dataset
sum(bc8_variants.df$total_af <= 0.01, na.rm=T)

min(bc8_variants.df$total_af, na.rm=T)
max(bc8_variants.df$total_af, na.rm=T) # no rare variants with MAF<0.01 and AF>0.99
sum(is.na(bc8_variants.df$total_af))

rare_in_wecare <- bc8_variants.df$total_af <= 0.01
sum(rare_in_wecare)

# Rare in controls
#rare_in_controls <- bc8_variants.df$controls_ac <= 0.05
#sum(rare_in_controls)

sets.ls <- list(gnomAD_NFE=bc8_variants.df$ID[rare_in_gnomAD_NFE],
            wecare=bc8_variants.df$ID[rare_in_wecare])

ggVennDiagram(sets.ls)

#rare <- rare_in_gnomAD_NFE | rare_in_wecare
rare <- rare_in_gnomAD_NFE & rare_in_wecare
#rare <- rare_in_gnomAD_NFE
#rare <- rare_in_controls

sum(rare)

#uncommon <- bc8_variants.df$vep_gnomAD_NFE_AF <= 0.05
#sum(uncommon)

# Clean-up
rm(rare_in_gnomAD_NFE, rare_in_wecare, sets.ls)

```

## Known benign or conflicting_interpretations

```{r}

# Known benign
known_benign <- 
  grepl("Benign",bc8_variants.df$CLNSIG) | 
  grepl("Likely_benign",bc8_variants.df$CLNSIG)

sum(known_benign)

conflicting_interpretations <- 
  grepl("Conflicting_interpretations",bc8_variants.df$CLNSIG)

sum(conflicting_interpretations)

```

## Combined functionaly important

```{r}
# Functionally important variants
fiv <- (predicted_lof | known_pathogenic | consensus_high_grade_missenses) & 
  rare & 
  !(conflicting_interpretations | known_benign)
sum(fiv)

bc8_fiv_all.df <- bc8_variants.df[fiv,] %>% 
  select(ID,vep_SYMBOL,vep_Consequence,vep_IMPACT,CLNSIG,CLNREVSTAT,PolyPhen_call,SIFT_call,
         vep_EXON,vep_gnomAD_NFE_AF,
         total_ac,total_an,total_af,
         controls_ac,controls_an,controls_af,
         cases_ac,cases_an,cases_af)
bc8_fiv_all.df
#View(bc8_fiv_all.df)

# Clean-up
rm(predicted_lof,known_pathogenic,rare,fiv,conflicting_interpretations,consensus_high_grade_missenses)

```

# Aggregated crude counts and Fisher test

```{r}

# Calculate crude counts
controls_ac <- sum(bc8_fiv_all.df$controls_ac) 
controls_an <- sum(bc8_fiv_all.df$controls_an)
controls_af <- controls_ac / controls_an

cases_ac <- sum(bc8_fiv_all.df$cases_ac)
cases_an <- sum(bc8_fiv_all.df$cases_an)
cases_af <- cases_ac / cases_an

# Calculate Fisher test
fisher_matrix <- matrix(c(cases_ac, cases_an - cases_ac,
                          controls_ac, controls_an - controls_ac),
                          nrow=2, byrow=T)

colnames(fisher_matrix) <- c("Alt","Ref")
rownames(fisher_matrix) <- c("Cases","Controls")
fisher_matrix 

ft <- fisher.test(fisher_matrix, conf.int=T)

ft

afs <- c(controls_af,cases_af)
names(afs) <- c("Controls","Cases")

cc <- table(phenotypes.df$cc)

mp <- barplot(afs, ylim=c(0,max(afs)+0.0005), 
main=paste("WECARE WES w/o Danish\n",cc[1],"controls vs ",cc[2]," cases\nCrude counts Fisher p=",round(ft$p.value,3)), ylab="Mean AF in selected variants")

text(mp,afs+0.0003,round(afs,4))

# Clean-up
rm(controls_ac, controls_an, controls_af,
   cases_ac, cases_an, cases_af,
   afs, mp, fisher_matrix, cc, ft)

```

# Carriers of the selected variants

```{r}

genotypes_bc8.mx <- genotypes.mx[bc8_fiv_all.df$ID,]
dim(genotypes_bc8.mx)

carriers.udf <- function(x){paste(names(x)[which(x>0)], collapse=",")}
x <- apply(genotypes_bc8.mx,1,carriers.udf)
x

rm(x, carriers.udf)

```

# Check data

```{r}

ls()

# Check objects dimentions
dim(bc8_variants.df)
dim(bc8_fiv_all.df)

dim(genotypes.mx)
dim(phenotypes.df)
dim(variants.df)

# Check that the tables and matrices are in sync
sum(phenotypes.df$wes_id != colnames(genotypes.mx))
sum(variants.df$ID != rownames(genotypes.mx))

```

# Save results

```{r}

save.image(file.path(data_folder,"s02_crude_counts_bc8_all.RData"))

```

# Final section

```{r}

sessionInfo()
Sys.time()
gc()

```
