---
title: "PCA"
author: "AL"
date: "04/10/2020"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

# Summary

Calculate PC-s for wecare wes dataset  

<style>
pre{
  overflow-x: auto;
}
pre code{
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}

options(width=999)

```

# Start section

```{r}

Sys.time()
gc()

rm(list=ls())

base_folder <- "/rds/project/erf33/rds-erf33-medgen//resources/1kgp/b38"
setwd(base_folder)

#install.packages("bigsnpr")
library(bigsnpr) # for bed_autoSVD() and bed()
library(bigutilsr) # for prob_dist() and tukey_mc_up()
library(ggplot2)
library(hexbin)

NCORES <- 2
#NCORES <- nb_cores() 

```

# Read data from bed-bim-fam file set

```{r}

bed_file <- file.path(base_folder,"data","s03_non_related","cclg.bed")

cclg.bed <- bed(bed_file) # bigsnpr::bed
cclg.bed

attributes(cclg.bed)
str(cclg.bed)

x <- bed_MAF(cclg.bed)
head(x)

cclg.bed$bedfile
cclg.bed$address

x <- cclg.bed$fam
class(x)
head(x)

x <- cclg.bed$map
class(x)
head(x)

cclg.maf <- bed_MAF(cclg.bed)
head(cclg.maf)

rm(x, bed_file)

```

# Calculate PCA

Takes care about LD etc.  
See ?plot.big_SVD for plotting svd objets.  

```{r}

# bigsnpr::bed_autoSVD, Default k = 10
cclg.svd <- bed_autoSVD(cclg.bed, ncores = NCORES) 

# Variants not in LD (detected by clumping during autoSVD)
vars_not_in_LD <- attr(cclg.svd, "subset")
length(vars_not_in_LD)

#attributes(cclg.svd)
str(cclg.svd)

# Eigenvalues
plot(cclg.svd$d)
plot(cclg.svd) # default type="screeplot" see ?plot.big_SVD  

# Eigenvectors
dim(cclg.svd$u)
head(cclg.svd$u)

plot(cclg.svd$u[,1],cclg.svd$u[,2])
plot(cclg.svd,type = "scores")
plot(cclg.svd,type = "scores",scores=1:10,coeff=0.4)

# Loadings
dim(cclg.svd$v)
head(cclg.svd$v)
plot(cclg.svd,type="loadings",loadings=1:10,coeff=0.4)

```

# Selecting outliers 

```{r}

# Eigenvectors
U <- cclg.svd$u

# Measure(s) of "outlieness"  
prob <- prob_dist(U, ncores=NCORES) # bigutilsr::prob_dist
S <- prob$dist.self / sqrt(prob$dist.nn) # alternatively: test$dist.self / test$dist.nn ?
tukey_threshold <- tukey_mc_up(S) # bigutilsr::tukey_mc_up

# Histogram by "outlieness" score
ggplot() +
  geom_histogram(aes(S), color = "black", fill = "blue", alpha = 0.3) +
  theme_bigstatsr() +
  geom_vline(xintercept=tukey_threshold, colour="red") +
  labs(x = "Statistic of outlierness (S)", y = "Frequency (sqrt-scale)")

# Indices for outliers/non-outliers (for convinience)
# Indices are numeric (not logical): rbigparallelr::rows_along(x): seq_len(nrow(x))
# The threshold of 0.9 has been selected manually from the hhistogram
samples_wo_outliers <- rows_along(cclg.bed)[S < 0.9]
outliers <- rows_along(cclg.bed)[S >= 0.9]

rm(tukey_threshold, U)

```

# Visualising "outlieness" on scatterplots

```{r}

# Scatterplot with "outlieness" score colourcoded
plot(cclg.svd, 
     type = "scores", 
     scores = c(1,2), 
     coeff = 0.6) +
  aes(color = S) + 
  scale_colour_viridis_c()

plot(cclg.svd, 
     type = "scores", 
     scores = c(1,2), 
     coeff = 0.6) +
  aes(color = S > 0.9)

```

# PCA w/o outliers

```{r}

# Repeating bed_autoSVD w/o outliers 
# and usisng already pre-selected variants.  
# To select the varaints, which are not in LD again: 
# ind.col=... and thr.r2=NA could be omitted.
cclg.no.outliers.svd <- bed_autoSVD(cclg.bed,
                        ind.row = samples_wo_outliers,
                        ind.col = vars_not_in_LD, 
                        thr.r2 = NA,
                        ncores = NCORES)

plot(cclg.no.outliers.svd)

plot(cclg.no.outliers.svd, 
     type = "scores", 
     scores = 1:10, 
     coeff = 0.4)

plot(cclg.no.outliers.svd, 
     type = "loadings", 
     loadings = 1:10, 
     coeff = 0.4)

```

# Project outliers 

A simple projecting procedure would be multiplying the genotypes by the corresponding PC loadings.  

However, the augmented algorithm (OADP) is used to avoid shrinkage (the shrincage is becoming evident in PCs > 10). 

```{r}

# Make NA matrix to allocate PCs for all samples: not-outliers and projected-outliers
PCs <- matrix(NA, nrow(cclg.bed), ncol(cclg.no.outliers.svd$u))

# A special case of using predict.big_SVD to copy
# scored available from cclg.no.outliers.svd to the new matrix
# (i.e. omitting the outliers rows)
PCs[samples_wo_outliers, ] <- predict(cclg.no.outliers.svd)

# Calculate the "projection" - whatever it is - for the outliers ...
proj <- bed_projectSelfPCA(cclg.no.outliers.svd, 
                           cclg.bed,
                           ind.row = outliers,
                           ncores = 1) # for a few individuals

# Add the "projected" scores for outlying indivoduals
PCs[outliers, ] <- proj$OADP_proj

# Plot projeted outliers
plot(PCs[samples_wo_outliers, 1:2], pch=1, xlab="PC1", ylab="PC2", 
     main="PC1 vs PC2 recalulated w/o outliers,\nthen outliers projected (red)")
points(PCs[outliers, 1:2], pch=19, col = "red")

```

# End section

```{r}

ls()
Sys.time()
gc()

```

